---
title: "segclust2d: bivariate segmentation with optional clustering for R"
output: github_document
---


```{r echo=FALSE, message=FALSE, results="hide"}
options(Encoding="UTF-8")
library(badger)
```


`r badger::badge_cran_release("segclust2d", "orange")`
`r badger::badge_cran_download("segclust2d", "grand-total", "yellow")`
`r badger::badge_devel("rpatin/segclust2d", "blue")`
`r badger::badge_last_commit("rpatin/segclust2d")`

# Introduction

`segclust2d` provides R code for two methods of segmentation and joint
segmentation/clustering of bivariate time-series. It was originally intended for
ecological segmentation (home-range and behavioural modes) but can be easily
applied on other type of time-series. The package also provides tools for
analysing outputs from R packages `moveHMM` and `marcher`.

# Website

Full documentation for segclust2d is available on this website:
[https://rpatin.github.io/segclust2d/](https://rpatin.github.io/segclust2d/)

Three topics are discussed there, and are also available as vignettes in the R package:


- [First, preparation of data to be analyzed with segmentation/clustering algorithm, including covariate calculations, subsampling and guidelines to prepare movement data.](https://rpatin.github.io/segclust2d/articles/v01_preparing_data.html)
- [Second, a guide to run the segmentation and segmentation/clustering method, including advise on setting parameters.](https://rpatin.github.io/segclust2d/articles/v02_run_segclust2d.html)
- [Finally, an overview of the possible outputs that can be generated with the package.](https://rpatin.github.io/segclust2d/articles/v03_explore_outputs.html)

# Installation

For the `r badger::badge_cran_release("segclust2d", "orange")` version  : 
```{r, eval=FALSE}
install.packages("segclust2d")
```

If you want the newest `r badger::badge_devel("rpatin/segclust2d", "blue")`, you can install `segclust2d` from github  with:

```{r, eval=FALSE}
devtools::install_github("rpatin/segclust2d")
```

