---
title: "3 - Exploring Outputs from segclust2d"
author: "R. Patin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3 - Exploring Outputs from segclust2d}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
```

```{r library and data, fig.show='hold'}
library(segclust2d)
```

# Possible outputs and general functioning

Both `segmentation()` and `segclust()` return objects of `segmentation-class` for which several function are available (see [below](#list-of-functions)). 

## General functioning

There are two types of functions: (1) some are general and show likelihood for all the different segmentation; (2) other are specific to a given segmentation and requires selecting a number of segment and of cluster (if applicable).

### Default values for nseg and ncluster

For the functions specific to a given segmentation, if you do not provide as argument the number of segment and of cluster, the functions will automatically select the best arguments based on a penalized log-likelihood as following:

- for outputs of `segmentation()` the optimal number of segments is selected with
[Lavielle`s criterium](https://rpatin.github.io/segclust2d/articles/run_segclust2d.html#selecting-the-number-of-segments-1). 
Other number may be provided with argument `nseg`.

- for outputs of `segclust()` the optimal number of cluster and segment is selected with a 
[BIC-based penalized criterium](https://rpatin.github.io/segclust2d/articles/run_segclust2d.html#selecting-the-number-of-segments-2). 
Other parameters may be provided with argument `nseg` and `ncluster`. It is recommended to manually choose the number of cluster based on biological knowledge or careful exploration of the BIC-based penalized likelihood, while using the BIC-selected number of segment for this number of cluster, by providing only argument `ncluster`.

### Graphical outputs

All plot methods use `ggplot2` package and return `ggplot` objects that can be further modified and customized using classical `ggplot2` (see [ggplot2 function reference](https://ggplot2.tidyverse.org/reference/).

### Default value for `order`

If you provide argument `order = TRUE` to a function specific to a segmentation, then the different segment or cluster will be numbered ordered by the variable provided as `order.var` in the `segmentation()` or `segclust()` call.

## List of functions

1. **Graphical outputs**

*For a specific segmentation:*

- `plot.segmentation` to show the time-series with the segment and if applicable clusters.
- `segmap` to show the results of the segmentation as a map of the trajectory (if applicable)
- `stateplot` plot summary statistics for all segment or clusters.

*Summary for all segmentations:*

- `plot_likelihood` for segmentation() show the log-likelihood of the segmentation for all numbers of segments.
- `plot_BIC` for segmentation() show the BIC-based penalized log-likelihood of the segmentation.clustering for all numbers of segments and cluster.


2. **Extracting results**

*For a specific segmentation:*

- `augment` returns a data.frame with the original data as well as the segment or cluster associated for each data point
- `segment` returns a data.frame with the beginning and end of each segment
- `states` for `segclust` provides a data.frame with summary statistics for all clusters

*Summary for all segmentations:*

- `logLik` for `segmentation()` returns a data.frame with the  log-likelihood for all number of segments. 
- `BIC` for `segclust()` returns a data.frame with the BIC-based penalized log-likelihood for all number of cluster and segments.

# Examples

As function are really similar between segmentation and segmentation/clustering, we will show examples mostly for the segmentation/clustering outputs, but the use is very similar, argument `ncluster` just need to be omitted for obtaining outputs for segmentation.

```{r loading data and segclust, fig.show='hold', message = FALSE}
data(simulmode)
simulmode$abs_spatial_angle <- abs(simulmode$spatial_angle)
simulmode <- simulmode[!is.na(simulmode$abs_spatial_angle), ]
mode_segclust <- segclust(simulmode,
                          Kmax = 20, lmin=10, ncluster = c(2,3),
                          seg.var = c("speed","abs_spatial_angle"),
                          scale.variable = TRUE)
```


## `plot.segmentation` for segmented time-series

```{r plot.segmentation, fig.show='hold', message = FALSE}
plot(mode_segclust, ncluster = 3)
```

## segmap - map the segmentation

`segmap()` plot the results of the segmentation as a map. This can be done only if data have a geographic meaning. Coordinate names are by default "x" and "y" but they can be provided through argument `coord.names`.

```{r segmap, fig.show='hold', message = FALSE, fig.width=5, fig.height=5}
segmap(mode_segclust, ncluster = 3)
```

## stateplot - plot states statistics

`stateplot()` show statistics for each state or segment.
```{r stateplot, fig.show='hold', message = FALSE, fig.width=3, fig.height=3}
stateplot(mode_segclust, ncluster = 3)
```

## Extract information from segmentation

### augment - get data.frame with segment/cluster information for all points

`augment.segmentation()` is a method for `broom::augment`. It returns an augmented data.frame with outputs of the model - here, the attribution to segment or cluster.

```{r augment, fig.show='hold', message = FALSE, eval = FALSE}
augment(mode_segclust, ncluster = 3)
```

### segment - Extract each segment (begin, end, statistics)

`segment()` allows retrieving informations on the different segment of a given segmentation. Each segment is associated with the mean and standard deviation for each variable, the state (equivalent to the segment number for `segmentation`) and the state ordered given a variable - by default the first variable given by `seg.var`. One can specify the variable for ordering states through the `order.var` of `segmentation()` and `segclust()`.

```{r segment, fig.show='hold', results = "hide", message = FALSE}
segment(mode_segclust, ncluster = 3)
```

### states - statistics about each states.

`states()` return information on the different states of the segmentation. For `segmentation()` it is quite similar to `segment()`. For `segclust`, however it gives the different cluster found and the statistics associated.

```{r states, fig.show='hold', results = "hide", message = FALSE}
states(mode_segclust, ncluster = 3)
```

## Get likelihood for all segmentation or segmentation/clustering

### log-Likelihood (segmentation)

`logLik.segmentation()` return information on the log-likelihood of the different segmentations possible. It returns a data.frame with the number of segment and the log-likelihood.

```{r simulshift, fig.show='hold', message = FALSE}
data("simulshift")
shift_seg <- segmentation(simulshift,
                          seg.var = c("x","y"),
                          lmin = 240, Kmax = 25,
                          subsample_by = 60)
```


```{r logLik, eval = FALSE}
logLik(shift_seg)

```

`plot_likelihood()` plot the log-likelihood of the segmentation for all the tested number of segments and clusters.

```{r plot_likelihood, fig.show='hold', message = FALSE}
plot_likelihood(shift_seg)
```

### BIC-based penalized likelihood (segclust)

`BIC.segmentation()` return information on the BIC-based penalized log-likelihood of the different segmentations possible. It returns a data.frame with the number of segment, the BIC-based penalized log-likelihood and the number of cluster. For `segclust()` only. Note that this does not truly returns a BIC. Here highest values are favored (in opposition to BIC)

```{r BIC, fig.show='hold', results = "hide", message = FALSE}
BIC(mode_segclust)
```

`plot_BIC()` plot the BIC-based penalized log-likelihood of the segmentation for all the tested number of segments and clusters.

```{r plot_BIC, fig.show='hold', message = FALSE}
plot_BIC(mode_segclust)
```


