---
title: "Using segTraj R package to segment trajectories"
author: "Rémi Patin - remi.patin@cefe.cnrs.fr"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes

vignette: >
  %\VignetteIndexEntry{Using segTraj R package to segment trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette present the use of `segTraj` R package that provides a method for segmenting trajectories acording to a bivariate algorithm, that is an extension of `adehabitatLT::lavielle`. This algorithm allow bivariate segmentation with a dynamic programming algorithm, along with the use of an EM (Expectation-Maximization) algorithm that allocates segment to different class. There are two different functions of segTraj package :
- `segTraj::DynProg` : segmentation only method using Dynamic Programming
- `segTraj::hybrid_simultanee` : clustering-segmentation alternating Dynamic Programming and EM.

The `segtools` package provides some tools to use easily the function from `segTraj`. This vignette is divided in two section, first a presentation of tools from `segTraj` package and then how to use them through `segtools` package.

```{r}
#devtools::install_github('MarieEtienne/segTraj')
#devtools::install_github('rpatin/segtools')

# library(segtools)
devtools::load_all()
```


## segTraj functions

`segTraj` can be installed through `devtools::install_github('MarieEtienne/segTraj')`

### Segmentation Only

`segTraj` functions takes as argument a two line matrix with each observation as column. Two functions are called for the segmentation-only algorithm : `Gmean_simultanee` and `DynProg`. Arguments are `lmin`, the minimum length of segments and `Kmax` the maximum number of segments.

```{r,cache=T}
data <- t(hrshift[,c("x","y")])
CostLoc <- segTraj::Gmean_simultanee(data, lmin = 100)
res.DynProg <- segTraj::DynProg(CostLoc, Kmax = 8)
```

After segmentation is done, we have to look at likelihood to see which model is the best. 3 appears to be the best choice, because adding a third segment still adds some likelihood, but a fourth one is not a strong gain.

```{r,fig.width = 6,fig.height=4}
plot(-res.DynProg$J.est,xlab='Number of segments',ylab='log-Likelihood',type='b')
abline(v=3,lty=2)
```

Now we can look at the segmentation.

```{r,fig.width = 8,fig.height=4,fig.cap = "The default plotting function of segTraj package"}
nseg=3
segTraj::bisig_plot(data,rupt = res.DynProg$t.est[nseg,])
```


### Clustering-Segmentation

Only `hybrid_simultanee` is called for joint clustering-segmentation algorithm.
Arguments are `lmin`, the minimum length of segments, `Kmax` the maximum number of segments, `P` the number of class and `sameSigma` deciding whether all segment have the same variance or not.

```{r,cache=T,results='hide',fig.keep='none'}
res.hybrid <- segTraj::hybrid_simultanee(data,P=3, Kmax=8,lmin=100, sameSigma = F)
```

After segmentation is done, we have to look at likelihood to see which model is the best. 3 appears to be the best choice, because adding more segment does not change much likelihood.

```{r,fig.width = 6,fig.height=4}
plot(res.hybrid$Linc,xlab='Number of segments',ylab='log-Likelihood',type='b')
abline(v=3,lty=2)
```

Now we can look at the segmentation.

```{r,fig.width = 8,fig.height=4,fig.cap = "The default plotting function of segTraj package"}
nseg=3
segTraj::bisig_plot(data,rupt = res.hybrid$param[[nseg]]$rupt,mu = res.hybrid$param[[nseg]]$phi$mu,pop = res.hybrid$param[[nseg]]$cluster)
```

## With `segtools` R package

`segtools` provides method for segmentation using `segTraj` functions and also usefull tools for plotting these segmentation. It can be installed through `devtools::install_github('rpatin/segtools')`

### Segmentation only

To create the segmentation we use `segtools::segmentation`. These creates an object of type `segmentation` for which several methods are available.

```{r,cache=T}

data.segmented <- segtools::segmentation(data = hrshift, seg.var = c("x","y"), order.var = "x", type = 'picard', picard.type = 'DynProg', lmin = 100, Kmax = 8)

```

To show likelihood, we use generic function `likelihood` that returns a ggplot object.

```{r,fig.width = 6,fig.height=4}
likelihood(data.segmented)
```

To show the segmentation we use the plot function for which we have to choose the number of segment, here 3 should again be the best.

```{r,fig.width = 8,fig.height=4}
nseg = 3
plot(data.segmented,nseg)
```

We can also show statistics for the different segment and plot their statistics.

```{r}
states(data.segmented,nseg = nseg)
```

In the case of a segmentation-only algorithm, each segment has a different state.

```{r,fig.width = 6,fig.height=4}
stateplot(data.segmented,nseg = nseg)
```

We can also show statistics on segments.

```{r}
segment(data.segmented,nseg = nseg)
```

Using `augment` we can extract the original data, augmented by a new column `state` that contains the estimated state/segment for a given number of segment.

```{r}
head(augment(data.segmented,nseg = nseg))
```

We can also display a map of the segmentation.

```{r,fig.width = 6,fig.height=6}
segmap(data.segmented,nseg = nseg)
```

### Clustering-Segmentation

To use the Clustering-Segmentation algorithm we just change `picard.type` to 'hybrid_simultanee' and specify the number of class.

```{r,cache=T,results='hide',fig.keep='none'}

data.segmented <- segtools::segmentation(data = hrshift, seg.var = c("x","y"), order.var = "x", type = 'picard', picard.type = 'hybrid_simultanee', lmin = 100, Kmax = 8,nclass=3)

```

To show likelihood, we use generic function `likelihood` that returns a ggplot object.

```{r,fig.width = 6,fig.height=4}
likelihood(data.segmented)
```

To show the segmentation we use the plot function for which we have to choose the number of segment, here 3 should again be the best.

```{r,fig.width = 8,fig.height=4}
nseg = 3
plot(data.segmented,nseg)
```

We can also show statistics for the different segment and plot their statistics.

```{r}
states(data.segmented,nseg = nseg)
```

In the case of a segmentation-only algorithm, each segment has a different state.

```{r,fig.width = 6,fig.height=4}
stateplot(data.segmented,nseg = nseg)
```

We can also show statistics on segments.

```{r}
segment(data.segmented,nseg = nseg)
```

Using `augment` we can extract the original data, augmented by a new column `state` that contains the estimated state/segment for a given number of segment.

```{r}
head(augment(data.segmented,nseg = nseg))
```

We can also display a map of the segmentation.

```{r,fig.width = 6,fig.height=6}
segmap(data.segmented,nseg = nseg)
```

